\large{\bf Looking forwards and backwards through locally regulated populations }

### 2023-03-10 forwards-and-backwards.tex
edits up to appendix

It is evident from the description of the process
(or, by differentiating in Definition~\ref{defn:lookdown_mgale})
that
\begin{align} \label{eqn:f_xi_mgale}
\begin{split}
    \langle f, \lp_t^N \rangle
    &=
    \langle f, \lp_0^N \rangle
    + M^f_t
\\ & \qquad {}
    +
    \int_0^t
    \big\langle
        c^N(x, \eta^N_s)
        \int_u^N \int_{\IR^d}
        \left(
            f(y, u_1) + f(x, u_1) + f(y, u) - f(x, u)
        \right)
        q^m(x, dy, \eta^N_s) du_1 
\\ & \qquad \qquad \qquad \qquad {}
        +
        \left(
            c^N(x, \eta^N_s) u^2
            - b^N(x, \eta^N_s) u
        \right)
        \frac{d}{du} f(x, u)
    ,
    \lp_s^N(dx, du) \bigg \rangle
    ds ,
\end{split}
\end{align}
where $M^f$ is a martingale
with quadratic variation
\begin{align} \label{eqn:f_xi_qv}
    \begin{split}
    [M^f]_t
    &=
    \int_0^t
    \bigg \langle c^N(x, \eta_s^N) \int_u^N \int_{\IR^d}
        \big[ f(y, u_1)^2
\\&\qquad \qquad {}
        + \left( f(x, u_1) + f(y, u) - f(x, u) \right)^2 \big]
    du_1 q^m(x, dy, \eta^N_s)
    ,
    \lp_s^N(dx, du) \bigg \rangle
    ds .
\end{split}\end{align}

@@

\begin{proof}[of Lemma \ref{lem:test_fn_tightness}]
    Again, we use the Aldous-Rebolledo criterion.
    Tightness of $\langle f, \lp_t \rangle$ for a fixed $t$
    follows from Lemma~\ref{lem:compact containment hn},
    so we need only prove conditions analogous to
    \eqref{eqn:eta_projections_goal1}
    and \eqref{eqn:eta_projections_goal2}
    applied to the martingale representation
    of equations~\eqref{eqn:f_xi_mgale} and~\eqref{eqn:f_xi_qv}.
    Indeed, the proof follows exactly as the proof of
    Lemma~\ref{lem:eta_projections_tightness},
    except that terms are bounded by constant multiples of $\langle h, \xi^N_s \rangle$
    instead of $\langle 1, \eta^N_s \rangle$.
    \comment{Give an example of this.}
    There is only one term that might use explanation,
    which is
    \begin{align*}
        &
        \frac{\theta}{N} \int_{\IR^d} \int_u^N (f(y, u) - f(x, u)) du_1 q^m(x, dy, \eta^N_s)
        \\ &\qquad
        % = \frac{\theta}{N} (N - u) \int_{\IR^d} (f(y, u) - f(x, u)) q^m(x, dy, \eta^N_s) \\ &\qquad
        = \left(1 - \frac{u}{N}\right) \theta \int_{\IR^d}
            (f(y, u) - f(x, u)) q^m(x, dy, \eta^N_s) ,
    \end{align*}
    but this is now bounded as in the proof of Lemma~\ref{lem:eta_f_bound}.
\end{proof}

@@

\comment{(TODO: Use notation for the motion of a lineage introduced in statement of theorem.)}

Here
\comment{(and in the statement of the theorem)}
we take the high-density, deterministic limit
(so, $\theta, N \to \infty$ and $\theta/N \to 0$)
and so suppose that the limiting process has density $\varphi_t(x)$
at location $x$ and time $t$.
Let $Y$ denote the spatial motion followed by a single line of descent.
Above equation~\eqref{eqn:limiting_generator},
we showed that $Y$ is a diffusion with generator at time $s$
$$
    \mathcal{L}^Y_s g(x) = \gamma ( \DG(rg)(x) - g \DG r(x) ) .
$$
The diffusion is time-inhomogeneous if the density is not constant with time
(since recall that the coefficients $\gamma$ and $r$ depend on the density).
Formally, the intensity of levels at $y$ at time $t$
that are descended from individuals that were at $x$ at time $s$
(with $s < t$) is
\begin{equation} \label{eqn:formal_intensity}
    \varphi_s(x) \IE_{s,x} \left[
        \exp\left(
            \int_s^t (F + \gamma \DG r)(Y_u) du
        \right)
        \ind_{Y_t = y}
    \right]
    dy ,
\end{equation}
where the subscript $s, x$ in the expectation indicates that $Y_s = x$.
To see why this should be true, 
suppose that an ancestor has level $v$. Conditional on its 
spatial motion $\{Y_u\}_{s\leq u\leq t}$, its level at time $t$ will
be $v \exp(-\int_s^t(F+\gamma\Delta r)(Y_u)du)$. This will be less than a given level 
$N$ if $v < N \exp(\int_s^t(F+\gamma\Delta r)(Y_u)du)$. 
The intensity of levels at $y$ that are descended from individuals at
$x$ can therefore be obtained as the limit as $N\to\infty$ of 
$1/N$ times the number of levels at $x$ at time zero with
$u<N \exp(\int_s^t(F+\gamma\Delta r)(Y_u)du)$ and for which
the corresponding individual is at $y$ at time $t$, which is 
precisely the quantity in~\eqref{eqn:formal_intensity}.

@@

To simplify to equation \eqref{eqn:lineage_generator2},
this, first define
$
    \DD f(x) = \sum_{ij} \covq_{ij} \partial_{ij} f(x),
$
% and note that $\DD$ satisfies the following identity:
% \begin{align}
%     \DD(fg)
%     &=
%     f \DD g
%     + 2 (\covq \grad f) \cdot \grad g
%     + g \DD f .
% \end{align}
and so the adjoint of $\DD$ is
$$
    \DD^* f(x)
    =
    \sum_{ij} \partial_{ij} (\covq_{ij} f(x)) .
$$
Note that $\DD^*$ satisfies the following identity:
\begin{align*}
    \DD^*(fg)
%     &=
%     \sum_{ij} \left\{
%         g \partial_{ij} (\covq_{ij} f)
%         + 2 \partial_{i} (\covq_{ij} f) \partial_j(g)
%         + \covq_{ij} f \partial_{ij} g
%     \right\} \\
    &=
    \sum_{ij} \left\{
        g \partial_{ij} (\covq_{ij} f)
        + 2 f \partial_{i} (\covq_{ij}) \partial_j(g)
        + 2 \covq_{ij} \partial_{i} (f) \partial_j(g)
        + \covq_{ij} f \partial_{ij} g
    \right\} \\
    &=
    g \DD^* f
    + 2 f \vec{c} \cdot \grad g
    + 2 (C \grad f) \cdot \grad g
    + f \DD g ,
\end{align*}
where $\vec{c}_j = \sum_i \partial_j C_{ij}$.
So, with $f = \gamma \varphi_s$,
\begin{eqnarray*}
\Lgen_s g
    &=&
    \frac{r}{\varphi_s} \left[
        \DD^*(\gamma \varphi_s g) - \grad \cdot (\gamma \varphi_s g \meanq)
        - g \DD^*(\gamma \varphi_s) + g \grad \cdot (\gamma \varphi_s \meanq)
    \right] \\
    &=&
    \frac{r}{\varphi_s} \left[
        \gamma \varphi_s \DD g
        + 2 \gamma \varphi_s \vec{c} \cdot \grad g
        + 2 (C \grad (\gamma \varphi_s)) \cdot \grad g
        - \gamma \varphi_s \meanq \cdot \grad g
    \right] \\
    &=&
    r \gamma \left[
        \DD g
        + 2 \vec{c} \cdot \grad g
        + 2 (C \grad \log(\gamma \varphi_s)) \cdot \grad g
        - \meanq \cdot \grad g
    \right] ,
\end{eqnarray*}
which is equation \eqref{eqn:lineage_generator}.

\begin{proof}[of Corollary \ref{cor:lineages_simple}]

In this section, we need only
simplify equation \eqref{eqn:lineage_generator}
in the case where $\covq = \sigma^2 I$.

    \comment{TODO}

\begin{proof}[of Corollary~\ref{cor:wavefront}
    \comment{
        TODO make non-redundant with proof of Theorem \ref{thm:lineages}.
    }

Let $Y$ denote the diffusion followed by following a single descendent line in this frame; 
i.e.~the diffusion with generator ${\cal L}^Y\phi=\gamma\Delta(r\phi)-\gamma\phi\Delta r -\mathfrak{c}\nabla\phi $. Note in particular here we take $\mathcal{B}=\Delta$. We start by expressing the evolution of a singles ancestry line in terms of the descendent line $Y$.

Formally, the intensity of levels at $y$ at time $t$ that are descended from 
individuals that were at $x$ at time $0$ is
\begin{equation}
\label{formal intensity [1]}
w(x)\IE_x\left[\exp\left(\int_0^t (F+\gamma \Delta r)(Y_s)ds\right)\ind_{Y_t=y}\right]dy,
\end{equation}
where the subscript $x$ in the expectation indicates that $Y_0=x$.
To see why this should be true, 
suppose that an ancestor has level $u$. Conditional on its 
spatial motion $\{Y_s\}_{0\leq s\leq t}$, its level at time $t$ will
be $u\exp(-\int_0^t(F+\gamma\Delta r)(Y_s)ds)$. This will be less than a given level 
$N$ if $u<N \exp(\int_0^t(F+\gamma\Delta r)(Y_s)ds)$. 
The intensity of levels at $y$ that are descended from individuals at
$x$ can therefore be obtained as the limit as $N\to\infty$ of 
$1/N$ times the number of levels at $x$ at time zero with
$u<N \exp(\int_0^t(F+\gamma\Delta r)(Y_s)ds)$ and for which
the corresponding individual is at $y$ at time $t$, which is 
precisely the quantity in~(\ref{formal intensity}).

It is easy to check that when we integrate~(\ref{formal intensity})
with respect to $x$ we recover $w(y)dy$. 
As a consequence of~(\ref{formal intensity}), writing $p(t,y,x)$ for
the probability density that the ancestor $t$ units in 
the past of an individual sampled 
at $y$ was at $x$, then, still formally,
\begin{equation}
\label{ptyx}
p(t,y,x)=\frac{w(x)}{w(y)}\IE_x\left[\exp\left(\int_0^t (F+\gamma\Delta r)(Y_s)ds\right)
\ind_{Y_t=y}\right].
\end{equation}

To make~(\ref{ptyx}) meaningful, we multiply by suitable test functions and 
integrate.
$$\int\int \psi(y)w(y)p(t,y,x)\phi(x)dydx=\int\phi(x)w(x)
\IE_x\left[\exp\left(\int_0^t(F+\gamma\Delta r)(Y_s)ds\right)\psi(Y_t)\right]dydx.$$

Writing $\hat{T}_t$ for the semigroup corresponding to the motion of 
ancestral lineages backwards in time (that is the semigroup corresponding
to $p(t,x,y)$), we can write this as 
$$\int \psi(y)w(y)\hat{T}_t\phi(y)dy=\int\phi(x)w(x)
\IE_x\left[\exp\left(\int_0^t(F+\gamma\Delta r)(Y_s)ds\right)\psi(Y_t)\right]dx.$$
Differentiating with respect to $t$ at $t=0$, and writing 
$\hat{\mathcal L}$ for the generator of $\hat{T}_t$, we find
\begin{eqnarray*}
\int \psi(y)w(y)\widehat{\mathcal L}\phi(y)dy&=&
\int\phi(x)w(x)\left({\mathcal L}^Y\psi(x)+\psi(x)(F+\gamma\Delta r)\psi(x)\right)dx\\
&=&\int\phi(x)w(x)\left(\gamma\Delta(r\psi)-\gamma\psi\Delta r -\mathfrak{c}\nabla\psi 
+(F+\gamma\Delta r)(\psi)\right)dx\\
&=&\int\phi(x)w(x)\left(\gamma\Delta(r\psi)-\mathfrak{c}\nabla\psi+F(\psi)\right)dx\\
&=&
\int\psi(x)\left[r\Delta(\gamma \phi w)+F\phi w+\mathfrak{c}\nabla(\phi w)\right]dx.
\end{eqnarray*}
Since $\psi$ was arbitrary,
\begin{eqnarray*}
\hat{\mathcal L}\phi&=&\frac{1}{w}\left[ r\Delta (\gamma\phi w)+F\phi w
+\mathfrak{c}\nabla(\phi w)\right]\\
&=&\frac{\phi}{w}\left[r\Delta(\gamma w)+Fw+\mathfrak{c}\nabla w\right]
+\frac{1}{w}\left[r\gamma w\Delta \phi +2\nabla (\gamma w)
\nabla\phi+\mathfrak{c}w\nabla\phi\right]\\
&=&r\gamma\left(\Delta\phi+2\frac{\nabla(\gamma w)}{\gamma w}\nabla\phi\right)
+\mathfrak{c}\nabla\phi,
\end{eqnarray*}
where we have used that $w$ solves~(\ref{profile}).

>>>>

It is evident from the description of the process
(or, by differentiating in Definition~\ref{defn:lookdown_mgale})
that
\begin{align} \label{eqn:f_xi_mgale}
\begin{split}
    \langle f, \lp_t^N \rangle
    &=
    \langle f, \lp_0^N \rangle
    + M^f_t
\\ & \qquad {}
    +
    \int_0^t
    \big\langle
        c_\theta(x, \eta^N_s)
        \int_u^N \int_{\IR^d}
        \left(
            f(y, u_1) + f(x, u_1) + f(y, u) - f(x, u)
        \right)
        q^m(x, dy, \eta^N_s) du_1 
\\ & \qquad \qquad \qquad \qquad {}
        +
        \left(
            c_\theta(x, \eta^N_s) u^2
            - b_\theta(x, \eta^N_s) u
        \right)
        \frac{d}{du} f(x, u)
    ,
    \lp_s^N(dx, du) \bigg \rangle
    ds ,
\end{split}
\end{align}
where $M^f$ is a martingale
with quadratic variation
\begin{align} \label{eqn:f_xi_qv}
    \begin{split}
    [M^f]_t
    &=
    \int_0^t
    \bigg \langle c_\theta(x, \eta_s^N) \int_u^N \int_{\IR^d}
        \big[ f(y, u_1)^2
\\&\qquad \qquad {}
        + \left( f(x, u_1) + f(y, u) - f(x, u) \right)^2 \big]
    du_1 q^m(x, dy, \eta^N_s)
    ,
    \lp_s^N(dx, du) \bigg \rangle
    ds .
\end{split}\end{align}

@@

\begin{proof}[of Lemma \ref{lem:test_fn_tightness}]
    Again, we use the Aldous-Rebolledo criterion.
    Tightness of $\langle f, \lp_t \rangle$ for a fixed $t$
    follows from Lemma~\ref{lem:compact containment hn},
    so we need only prove conditions analogous to
    \eqref{eqn:eta_projections_goal1}
    and \eqref{eqn:eta_projections_goal2}
    applied to the martingale representation
    of equations~\eqref{eqn:f_xi_mgale} and~\eqref{eqn:f_xi_qv}.
    Rewriting~\eqref{eqn:f_xi_mgale} with $c_\theta = c_\theta(x, \eta_s)$,
    \begin{align*}
        \langle f, \lp_t \rangle
        &=
        \langle f, \lp_0 \rangle
        + M^f_t
        + \int_0^t \Big\langle
            c_\theta \int_u^N \int (f(y, u_1) + f(x, u_1)) q^m(x, dy) du_1
        \\ {} &\qquad
            + c_\theta (N - u) \int (f(y, u) - f(x, u)) q^m(x, dy)
            + (c_\theta u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle ds
        &\le
        \langle f, \lp_0 \rangle
        + M^f_t
        + \int_0^t \Big\langle
            c_\theta \int_u^N \int (f(y, u_1) + f(x, u_1)) q^m(x, dy) du_1
        \\ {} &\qquad
            + c_\theta (N - u) \int (f(y, u) - f(x, u)) q^m(x, dy)
            + (c_\theta u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle ds
    \end{align*}
    The bounds analogous to \eqref{eqn:eta_projections_goal1} and \eqref{eqn:eta_projections_goal2}
    follow as in the proof of Lemma~\ref{lem:eta_projections_tightness}:
    for instance, observe that using that $c_\theta \le C_0$ for some $C_0$,
    the predictable part of this semimartingale decomposition is bounded by
    \begin{align*}
        \Big\langle
            2 C_0 \|f\|_\infty u_f
            + (1 - u/N) \gamma B_f
            + (C_0 u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle
        &\le
        \langle
            C_0 u_f^2 + \sup_x |b_\theta(x, \eta_s)| u_f \|\frac{d}{du} f\|_\infty
        \rangle ,
    \end{align*}
    which can be bounded as we did for \eqref{eqn:eta_projections_goal1}.
\end{proof}

@@

Here we have taken the high-density, deterministic limit
(so, $\theta, N \to \infty$ and $\theta/N \to 0$)
and so suppose that the limiting process has density $\varphi_t(x)$
at location $x$ and time $t$.
Let $Y$ denote the spatial motion followed by a single line of descent.
Above equation~\eqref{eqn:limiting_generator},
we showed that $Y$ is a diffusion with generator at time $s$
$$
    \mathcal{L}^Y_s g(x) = \gamma ( \DG(r(\cdot,\eta_s) g(\cdot))(x) - g(x) \DG r(x,\eta_s) ) .
$$
% (This is $\mathcal{L}^Y_s g(x) = \gamma ( \DG(rg)(x) - g \DG r(x) )$ in shorter but more cryptic notation.)
The diffusion is time-inhomogeneous if the density is not constant with time
(since recall that the coefficients $\gamma$ and $r$ depend on the density).
Let $\varphi_t(x)$ be the limiting density, which solves \eqref{general deterministic limit},
$\partial_t \varphi = r \DG^*[ \varphi_t \gamma ] + \varphi F$.
Formally, the intensity of levels at $y$ at time $t$
that are descended from individuals that were at $x$ at time $s$
(with $s < t$) is
\begin{equation} \label{eqn:formal_intensity}
    \varphi_s(x) \IE_{s,x} \left[
        \exp\left(
            \int_s^t (F + \gamma \DG r)(Y_u) du
        \right)
        \ind_{Y_t = y}
    \right]
    dy ,
\end{equation}
where the subscript $s, x$ in the expectation indicates that $Y_s = x$.
To see why this should be true, 
suppose that an ancestor at time $s$ has level $v$. Conditional on its 
spatial motion $\{Y_u\}_{s\leq u\leq t}$, its level at time $t$ will
be $v \exp(-\int_s^t(F+\gamma\Delta r)(Y_u)du)$. This will be less than a given level 
$\lambda$ if $v < \lambda \exp(\int_s^t(F+\gamma\Delta r)(Y_u)du)$. 
The intensity of levels at $y$ that are descended from individuals at
$x$ can therefore be obtained as the limit as $\lambda\to\infty$ of 
$1/\lambda$ times the number of levels at $x$ at time zero with
$u<\lambda \exp(\int_s^t(F+\gamma\Delta r)(Y_u)du)$ and for which
the corresponding individual is at $y$ at time $t$, which is 
precisely the quantity in~\eqref{eqn:formal_intensity}.

@@

To simplify to equation \eqref{eqn:lineage_generator2},
first define
$
    \DD f(x) = \sum_{ij} \covq_{ij} \partial_{ij} f(x),
$
% and note that $\DD$ satisfies the following identity:
% \begin{align}
%     \DD(fg)
%     &=
%     f \DD g
%     + 2 (\covq \grad f) \cdot \grad g
%     + g \DD f .
% \end{align}
and so the adjoint of $\DD$ is
$$
    \DD^* f(x)
    =
    \sum_{ij} \partial_{ij} (\covq_{ij} f(x)) .
$$
Note that $\DD^*$ satisfies the following identity:
\begin{align*}
    \DD^*(fg)
%     &=
%     \sum_{ij} \left\{
%         g \partial_{ij} (\covq_{ij} f)
%         + 2 \partial_{i} (\covq_{ij} f) \partial_j(g)
%         + \covq_{ij} f \partial_{ij} g
%     \right\} \\
    &=
    \sum_{ij} \left\{
        g \partial_{ij} (\covq_{ij} f)
        + 2 f \partial_{i} (\covq_{ij}) \partial_j(g)
        + 2 \covq_{ij} \partial_{i} (f) \partial_j(g)
        + \covq_{ij} f \partial_{ij} g
    \right\} \\
    &=
    g \DD^* f
    + 2 f \vec{c} \cdot \grad g
    + 2 (C \grad f) \cdot \grad g
    + f \DD g ,
\end{align*}
where $\vec{c}_j = \sum_i \partial_j C_{ij}$.
So, with $f = \gamma \varphi_s$,
\begin{eqnarray*}
\Lgen_s g
    &=&
    \frac{r}{\varphi_s} \left[
        \DD^*(\gamma \varphi_s g) - \grad \cdot (\gamma \varphi_s g \meanq)
        - g \DD^*(\gamma \varphi_s) + g \grad \cdot (\gamma \varphi_s \meanq)
    \right] \\
    &=&
    \frac{r}{\varphi_s} \left[
        \gamma \varphi_s \DD g
        + 2 \gamma \varphi_s \vec{c} \cdot \grad g
        + 2 (C \grad (\gamma \varphi_s)) \cdot \grad g
        - \gamma \varphi_s \meanq \cdot \grad g
    \right] \\
    &=&
    r \gamma \left[
        \DD g
        + 2 \vec{c} \cdot \grad g
        + 2 (C \grad \log(\gamma \varphi_s)) \cdot \grad g
        - \meanq \cdot \grad g
    \right] ,
\end{eqnarray*}
which is equation \eqref{eqn:lineage_generator2},
\end{proof}
